---
title: "Chen_Final Project"
author: "Emma Chen"
date: "11/21/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir = 'C:/Users/Emma/Documents/BMIN503_Final_Project')
getwd()
```

### Overview
My project study whether administration of both JAK inhibitors and immune checkpoint blockade therapy (anti-PD1) improves tumor resistance so patients will response to immunotherapy better, improve survival and have more powerful tumor response. My dataset included Flow Cytometry and CyTOF of 1 of 7 patients’ PBMC blood sample across 8 timepoints (because of the computing power of computer, I was not able to analyze across multiple patient data.

### Introduction 
My project is to investigate whether combined administration of JAK inhibitor and immune checkpoint blockade (ICB)–anti-PD1 immunotherapy is better than with only ICB over time. Currently, ICB, either anti-PD1 or anti-CTLA4, is used to treat cancer patients. ICB functions as taking the brake off immune system’s response to cancer. However, many patients treated with ICB relapsed or their cancer progresses, leading to the idea that taking one break off is not enough, multiple brakes needs to be released. Dr. Minn at Penn Medicine found out that interferon (IFN) pathway is critical to tumor’s resistance to immunotherapy, which also means shutting down the IFN pathway may improve the response to ICB. He found out that Janus kinase (JAK) inhibitor could overcome the resistance to immunotherapy, improving survival and powerful tumor responses.

Since this finding, some patients were given only ICB or both ICB and JAK inhibitors to test if JAK inhibitors actually improve tumor resistance problem. Clinical data and blood samples were obtain from 7 patients in our study across 8 different timepoints (one treatment cycle every 3 weeks). Blood samples were processed to PBMC, stained with 25 extracellular and intracellular protein markers in total, ran in both Flow Cytometry and CyTOF (using metal ions instead of fluorochromes) to assess immune profile. There are multiple parameters/variables in the diverse datasets that worth investigating.

There are 4 objectives in total.
-First is to use different R packages to analyze both flow and mass cytometry data; current analysis workflow is to use FlowJo or CyTOF to analyze cytometry data. But as manual gating is time consuming for so many markers and very subjective according to the person doing analysis, we decided to explore if R packages could finish some steps in analysis workflow.
-Second is to design and simplify (longitudinal) data analysis workflow for different dataset type, different study goals, different dataset size, variables, etc.
-Third is so compare Immune Checkpoint Blockage (ICB) only versus both JAK inhibitors and ICB combination, allowing us to confirm main hypothesis whether IFN pathway inhibition improve ICB immunotherapy effects. -Last, to pipeline analyzing multiple patients' data.

Last but not least, clinical data can be used to profile patients and determine if clinical data has correlation to patient’s immune profile.

### Methods
Data analysis workflow for cytometry data have a few steps:
-concatenate multiple files into one file (which files to concatenate depends on study questions). In my project, I concatenate all 8 timepoints' .fcs files of 1 patient together (will indicate below).
-input .fcs files into readable format in R, such as matrixs.
-clean up matrix:define which column has which protein markers, choose protein markers of interest, delete columns would not use for following analysis, such as FSC, SSC gates based on cell size and granularity, or dead cells and dump (cells other than target cell types).
-reduce dimensionality. Here we evaluate two packages: viSNE and UMAP based on how well they show large-scale information, data-consistency and run time.
-cluster cells based on common markers they expressed, also known as their fingerprints.
-use MEM package (Marker Enrichment Modeling, developed by the Irish Lab of Vanderbilt University) to do immune profiling and dig deeper into cluster features 


### Results
Step 0.1: run libraries needed for analysis
```{r}
library(FlowSOM)
library(flowCore)
library(Biobase)
library(gplots)
library(ggplot2)
library(hexbin)
library(MEM)
library(tidyverse)
library(Rtsne)
library(uwot)
library(viridis)
library(ggExtra)
library(premessa)
```

Step 0.2: set and confirm working directory
```{r}
setwd("C:/Users/Emma/Documents/BMIN503_Final_Project")
getwd()
```

Step 1: use package Premessa 's concatenate_fcs_files function to concatenate files. Out put as .fcs
```{r}
concat <-  dir(pattern="*.fcs")
concatenate_fcs_files(concat, output.file = "C:/Users/Emma/Documents/BMIN503_Final_Project/concatnated.fcs")
```

Step 2: Cleanup data for analysis:
look at concatenated fcs file first to find out markers names.
replace with fcs file name of dataset
Read in concatenated FCS file
```{r}
test <-  dir(pattern="concatnated.fcs")


test.data = read.FCS(test, transformation="linearize", which.lines=NULL,
         alter.names=FALSE, column.pattern=NULL, invert.pattern = FALSE,
         decades=0, ncdf = FALSE, min.limit=NULL, 
         truncate_max_range = TRUE, dataset=NULL, emptyValue=TRUE, 
         channel_alias = NULL)

colnames(test.data)
markernames(test.data)
```
columns need to be renamed: 7:31
columns need to be used for SOM
There are 25 markers for later analysis: "CD45RA, CXCR5, PD1, IgG4, dump, CD8, CD127, Tox, Tim3, Eomes, TCF1, CD38, CCR7, CTLA-4, CD226, CD39, FoxP3, Ki67, CD27, CD4, CD73,TIGIT, T-bet, CD28, CD3"

Step 3.1: convert .fcs to flowFrame and to Matrix, define columns of protein markers of interest as previously discussed.
```{r}
all.data = exprs(test.data)
scaled.data = all.data[, c(7:31)]
overall_seed = 46
```


Step 3.2a Run t-SNE on markers of interest, reduce dimension
```{r}
set.seed(overall_seed)
mytSNE = Rtsne(
  scaled.data, #column 7:31
  dims = 2,
  initial_dims = 8,
  perplexity = 15,
  check_duplicates = FALSE,
  max_iter = 10000
)
tSNE.data = as.data.frame(mytSNE$Y) 
#Matrix containing the new representations for the objects

#open mytSNE to find out what is range
range <- apply(apply(tSNE.data, 2, range), 2, diff)
graphical.ratio.t <- (range[1] / range[2])


# t-SNE flat dot plot and density dot plot (each dot is a patient)
tSNE.plot <- data.frame(x = tSNE.data[, c(1)], y = tSNE.data[, c(2)])

ggplot(tSNE.plot) + coord_fixed(ratio = graphical.ratio.t) + 
  geom_point(aes(x = x, y = y), cex = 1) + 
  labs(x = "t-SNE 1", y = "t-SNE 2", title = "t-SNE on 1 patient's 8 timeponts data") + 
  theme_bw() + 
  labs(caption = "Data from Divij Mathew/Wherry Lab")

```

note: 1 file with 8000 events, size 1.5Mb, took around 10mins to run 
(will compare to UMAP later)


Step 4: cluster tSNE plot with flowSOM, color code clusters. clusters = 20 based on previous manual gating.
```{r}
# Run FlowSOM on the t-SNE axes
tSNE.data.mat <- as.matrix(tSNE.data)

# create flowFrame
metadata <-
  data.frame(name = dimnames(tSNE.data.mat)[[2]],
             desc = paste('t-SNE', dimnames(tSNE.data.mat)[[2]]))
metadata$range <- apply(apply(tSNE.data.mat , 2, range), 2, diff)
metadata$minRange <- apply(tSNE.data.mat , 2, min)
metadata$maxRange <- apply(tSNE.data.mat , 2, max)
tSNE.flowframe <- new("flowFrame",
                      exprs = tSNE.data.mat ,
                      parameters = AnnotatedDataFrame(metadata))

# implement the FlowSOM on t-SNE data
fSOM.t <-
  FlowSOM(
    tSNE.flowframe,
    compensate = FALSE,
    transform = FALSE,
    toTransform = c(1:2),
    scale = TRUE,
    colsToUse = c(1:2),
    xdim = 10,
    ydim = 10,
    nClus = 20, #change if needed
    seed = overall_seed
  )
tSNE.FlowSOM.clusters <-
  as.matrix(fSOM.t[[2]][fSOM.t[[1]]$map$mapping[, 1]])

# plot t-SNE with FlowSOM clusters
ggplot(tSNE.plot) + coord_fixed(ratio = graphical.ratio.t) + 
  geom_point(aes(x = x, y = y, color = tSNE.FlowSOM.clusters), cex = 1.5) + 
  labs(x = "t-SNE 1", y = "t-SNE 2", title = "FlowSOM Clustering on t-SNE Axes", 
       color = "FlowSOM Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=5))) +
  labs(caption = "Data from Divij Mathew/Wherry Lab")
```

step 5: immune profile each cluster with MEM. MEM is good at look at markers that do NOT have a big variation across cells.
```{r}
# Run MEM on the FlowSOM clusters from t-SNE
cluster = as.numeric(as.vector((tSNE.FlowSOM.clusters)))
MEMdata = cbind(scaled.data, cluster)

MEM.values.tf = MEM(
  MEMdata,
  transform = TRUE,
  cofactor = 150,
  choose.markers = FALSE,
  markers = "all",
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD45RA, CXCR5, PD1, IgG4, dump, CD8, CD127, Tox, Tim3, Eomes, TCF1, CD38, CCR7, CTLA-4, CD226, CD39, FoxP3, Ki67, CD27, CD4, CD73,TIGIT, T-bet, CD28, CD3",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL,
  output.prescaled.MEM = TRUE
)


# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.tf,
  cluster.MEM = "both",
  cluster.medians = "none",
  display.thresh = 1,
  newWindow.heatmaps = FALSE,
  output.files = TRUE,
  labels = TRUE,
  only.MEMheatmap = FALSE
)
```

Now let's look at how UMAP reduce dimensionality of the SAME dataset for visualization.

Step 3.2b: using UMAP to reduce dimensionality
```{r}
# prepare data for use in UMAP
data <- lapply(lapply(test, read.FCS), exprs) #list apply to read FCS file (read.FCS) and to make it readable (exprs), return list for UMAP. apply takes in 3 arguments: 1 data matrix, 2 row/column operation, function to be applied 
ID <- c(1:length(data)) #ID for each file as files are in object (data) now
combined.data = as.data.frame(do.call(rbind, mapply(
  cbind, data, "File ID" = ID, SIMPLIFY = F
)))
combined.data$`File ID` <- as.numeric(combined.data$`File ID`)
chosen.markers = combined.data[, c(7:31)]
transformed.chosen.markers <- chosen.markers %>%
  mutate_all(function(x)
    asinh(x / 15))
overall_seed = 43 
```


```{r UMAP}
# Run UMAP on all surface markers
set.seed(overall_seed)
myumap <-
  umap(transformed.chosen.markers,
       ret_model = TRUE,
       n_threads = 1, 
       verbose = TRUE)
umap.data = as.data.frame(myumap$embedding)

range <- apply(apply(umap.data, 2, range), 2, diff)
graphical.ratio <- (range[1] / range[2])

# UMAP flat dot plot and density dot plot
UMAP.plot <- data.frame(x = umap.data[, 1], y = umap.data[, 2])

ggplot(UMAP.plot) + coord_fixed(ratio = graphical.ratio) + 
  geom_point(aes(x = x, y = y), cex = 1) + labs(x = "UMAP 1", y = "UMAP 2", 
                                                title = "UMAP on 1 patient's 8 timepoints Data") + 
  theme_bw() + 
  labs(caption = "Data from Divij Mathew/Wherry Lab")

ggplot(UMAP.plot, aes(x = x, y = y)) + coord_fixed(ratio = graphical.ratio)  + 
  geom_bin2d(bins = 128) +
  scale_fill_viridis_c(option = "A", trans = "sqrt") + 
  scale_x_continuous(expand = c(0.1, 0)) +
  scale_y_continuous(expand = c(0.1, 0)) + labs(x = "UMAP 1", y = "UMAP 2", 
                                                title = "UMAP on 1 patient's 8 timepoints Data") + 
  theme_bw() + 
  labs(caption = "Data from Divij Mathew/Wherry Lab")
```
Note: same file with 8000 events, size 1.5Mb took 1mins to run a UMAP map.UMAP has timestamp 
One advantage of UMAP over tSNE as a dimension reduction tool is that UMAP has shorter run time. 



Step 4: cluster UMAP plot with flowSOM, color code clusters. clusters = 20 based on previous manual gating.
```{r FlowSOM on UMAP}
# Run FlowSOM on the UMAP axes
umap.matrix <- as.matrix(umap.data)

# create flowFrame
UMAP.metadata <-
  data.frame(name = dimnames(umap.matrix)[[2]],
             desc = paste('UMAP', dimnames(umap.matrix)[[2]]))
UMAP.metadata$range <- apply(apply(umap.matrix, 2, range), 2, diff)
UMAP.metadata$minRange <- apply(umap.matrix, 2, min)
UMAP.metadata$maxRange <- apply(umap.matrix, 2, max)
umap.flowframe <- new("flowFrame",
                      exprs = umap.matrix,
                      parameters = AnnotatedDataFrame(UMAP.metadata))

# implement the FlowSOM on the data
fsom <-
  FlowSOM(
    umap.flowframe,
    compensate = FALSE,
    transform = FALSE,
    toTransform = c(1:2),
    scale = TRUE,
    colsToUse = c(1:2),
    nClus = 20, #change this
    seed = overall_seed
  )
FlowSOM.clusters <-
  as.matrix(fsom[[2]][fsom[[1]]$map$mapping[, 1]])

# plot FlowSOM clusters on UMAP axes
ggplot(UMAP.plot) + coord_fixed(ratio=graphical.ratio) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters),cex = 1.5) + 
  labs(x = "UMAP 1", y = "UMAP 2",title = "FlowSOM Clustering on UMAP Axes", 
       color = "FlowSOM Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=5)))+
  labs(caption = "Data from Divij Mathew/Wherry Lab")
```
Note: 
comparing this dimension reduction graph from UMAP, we can see more distinct clusters than tSNE's.
Another advantage of UMAP over tSNE is that it can show inter-cluster relation better, which is very important if dataset size increases. 

step 5: immune profile each cluster with MEM. MEM is good at look at markers that do NOT have a big variation across cells.
```{r MEM on UMAP/FlowSOM Clusters}
# Run MEM on the FlowSOM clusters from UMAP
cluster = as.numeric(as.vector((FlowSOM.clusters))) #clusters from UMAP and FlowSOM clustering
MEM.data = cbind(transformed.chosen.markers, cluster)

MEM.values.uf = MEM(
  MEM.data,
  transform = FALSE,
  cofactor = 150,
  choose.markers = FALSE,
  markers = "all",
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD45RA, CXCR5, PD1, IgG4, dump, CD8, CD127, Tox, Tim3, Eomes, TCF1, CD38, CCR7, CTLA-4, CD226, CD39, FoxP3, Ki67, CD27, CD4, CD73,TIGIT, T-bet, CD28, CD3",
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL, 
  output.prescaled.MEM = TRUE
)

# build MEM heatmap and output enrichment scores
build.heatmaps(
  MEM.values.uf,
  cluster.MEM = "both",
  cluster.medians = "none",
  display.thresh = 1,
  newWindow.heatmaps = FALSE,
  output.files = TRUE,
  labels = TRUE,
  only.MEMheatmap = FALSE
)
```

RMSD calculate pair-wise cluster relationship. It determines which markers the population share in common and then calculates the percent maximum RMSD between the two populations. the more red means the two clusters are more similar. With this pair-wise comparison for different population, we can then look into the two cluster that are similar, from UMAP and from tSNE, look at their immune profile. 

future directions: compare clusters from different timepoints of the same patients, and between patients. the more red means the clusters have more enrichment MEM scores in common. 
```{r RMSD for All Clusters}
# RMSD to compare labels from all populations, cluster from TF vs clusters
tf.MEM.scores = as.data.frame(MEM.values.tf[[5]])
rownames(tf.MEM.scores) = paste0(rownames(tf.MEM.scores), ' (TF)')
uf.MEM.scores = as.data.frame(MEM.values.uf[[5]])
rownames(uf.MEM.scores) = paste0(rownames(uf.MEM.scores), ' (UMAP)')
all.MEM.values = as.matrix(rbind(uf.MEM.scores, tf.MEM.scores))

RMSD_vals <-
  MEM_RMSD(
    all.MEM.values,
    format = NULL,
    newWindow.heatmaps = FALSE,
    output.matrix = FALSE
  )
```

Now we have 20 clusters from either tSNE+flowSOM's MEM, or from UMAP+flowSOME's MEM, then we can look at how these clusters express different proteins using machine learning. 

1st: PCA

Cleaning data. Import from txt generate from MEM in the output files' folder. Add a column containing clusterIDs (1-20) for our reference and analysis.

first we can look at tSNE+flowSOM+MEM clusters
```{r}
library(ggbiplot)

t <- read.table(file = "C:/Users/Emma/Documents/BMIN503_Final_Project/output files/2019-12-13_161409 Pre-scaled MEM matrix.txt", #Name of file
                header = TRUE, #Does the file have a header (names of columns)?
                sep = "\t", #Character that separates  column entries (tab, space, comma)
                skip = 0, #Number of lines to skip from beginning
                as.is = TRUE, #Should character variables be left alone 
                na.strings = NA, #The encoding of missing values. Blanks are assumed missing too.
                comment.char = "#", #Lines of file beginning with this character are ignored. (# default)
                quote = "", #Character used to quote values. "" is no quotes.
                nrows = -1 #The number of rows to read in. Negative/invalid values are ignored
)

cluster <- unlist(rownames(t))
t <- t %>%
  mutate(clusterID = cluster)
```

PCA analysis of 20 clusters from UMAP+flowSOM' MEM. 
```{r}
tMEM.pca <- prcomp(t[, 1:25], scale = TRUE)
summary(tMEM.pca)

ggbiplot(tMEM.pca, groups = t$clusterID, circle = TRUE) +
  scale_color_discrete(name = '') +
  theme(legend.direction = 'horizontal', legend.position = 'top')

ggscreeplot(tMEM.pca)
```

Next look at UMAP+flowSOM+MEM clusters.
```{r}
u <- read.table(file = "C:/Users/Emma/Documents/BMIN503_Final_Project/output files/2019-12-13_161441 Pre-scaled MEM matrix.txt", #Name of file
                header = TRUE, #Does the file have a header (names of columns)?
                sep = "\t", #Character that separates  column entries (tab, space, comma)
                skip = 0, #Number of lines to skip from beginning
                as.is = TRUE, #Should character variables be left alone 
                na.strings = NA, #The encoding of missing values. Blanks are assumed missing too.
                comment.char = "#", #Lines of file beginning with this character are ignored. (# default)
                quote = "", #Character used to quote values. "" is no quotes.
                nrows = -1 #The number of rows to read in. Negative/invalid values are ignored
)

cluster2 <- unlist(rownames(u))
u <- u %>%
  mutate(clusterID = cluster2)
```

PCA analysis of 20 clusters from UMAP+flowSOM' MEM. 
```{r}
uMEM.pca <- prcomp(u[, 1:25], scale = TRUE)
summary(uMEM.pca)

ggbiplot(uMEM.pca, groups = u$clusterID, circle = TRUE) +
  scale_color_discrete(name = '') +
  theme(legend.direction = 'horizontal', legend.position = 'top')

ggscreeplot(uMEM.pca)
```
Notes: comparing PCA of UMAP and tSNE clusters, cluster's immune profiles and distribution is quite different. 



###Discussion: 
The two previous methods use viSNE and UMAP to reduce dataset dimension. 

There are a few limitations of viSNE: 
1-It could not meaningfully represent very large datasets
	it might lose large-scale info, the inter-cluster relationship, 
	this could be a problem when we are analyzing genomic data, 
	it does not scale really well for rapidly increasing sample sizes in scRNA seq.
2-and it takes a longer time to compute (my project's viSNE took more than 15mins to run on 8000 events, UMAP took less than 1min, with start and end timestamp)

On the other hand, UMAP has a faster runtime, data consistency and more meaningful organization of cell clusters
That does not mean viSNE is inferior to UMAP. To use which data visualization methods, PCA, viSNE, UMAP or others, depends on what questions we are trying to answer, the dataset size, variability.  

As for designing cytometry workflow, R is capable to do concatenation, gating based on markers threshold, reduce dimensionality without GUI, cluster without GUI, and do immume profiling even protein marker does not vary obviously in manual gating. 

But initial gating should better be done for cells of interest. 

Completed tasks are: 
1)Alternative and potentially more efficient way to analyze cytometry data (Flow & CyTOF)
2)Create workflow for analyzing longitudinal study data 
3)Evaluated dimension reduction tool/packages
-Depends on study goals, data variation, dataset size, etc.
4)look at 20 clusters' protein expression with PCA 

###Future directions: 
1-Immune profile changes with combined therapy vs only ICB
  -Differentiate each timepoint from concatenated file
  -Investigate how cluster changes according to timepoints

2–increase dataset size to study multiple patients
  -Analyze multiple patients at the same time
  -Evaluate if JAK inhibitor resolves tumor resistance to ICB from larger dataset

3–analyze clinical data and experiment data side by side




Github update command:
github update
git add -A
git commit -m "[message about commit]"
git push
git status
